---
title: "CLINICAL STUDY OF KIDNEY DISEASE"
author: "Enock Bereka"
date: "2025-02-22"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Load the necessary libraries
```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(performance)
library(ggeffects)
library(ggcorrplot)
library(equatiomatic)
library(sjPlot)
library(gtsummary)
library(report)
library(vip)
library(car)
library(naniar)
library(gridExtra)
library(caret)
```

## Load and inspect the dataset
```{r warning=FALSE, message=FALSE}
data <- read_csv("C:/Users/PC/OneDrive/Desktop/Data Science/Datasets/Datasets/kidney_disease.csv")
glimpse(data)
data.frame(colSums(is.na(data)))
sum(duplicated(data))
```

## Data preparation
## Handle missing values by replacing them with the
median (for numeric) or mode (for categorical)
```{r warning=FALSE, message=FALSE}
data$rc <- as.factor(data$rc)
data$rc <- as.numeric(data$rc)
data <- data %>%
  mutate(across(where(is.numeric), ~replace_na(., median(., na.rm=TRUE)))) %>%
  mutate(across(where(is.character), ~replace_na(., as.factor(mode(na.omit(.))))))
```

## Label encoding
```{r warning=FALSE, message=FALSE}
data <- data %>% mutate_if(is.character, as.factor)
```

## Rename the variables
```{r warning=FALSE, message=FALSE}
data <- rename(data,
               "id" = id,
               "Blood_pressure" = bp,
               "Specific_graphity" = sg,
               "Albumin" = al,
               "Sugar" = su,
               "Red_blood_cells" = rbc,
               "Pus_cell" = pc,
               "pus_cell_clamps" = pcc,
               "Bacteria" = ba,
               "Blood_glucose_random" = bgr,
               "Blood_urea" = bu,
               "Serum_creatanine" = sc,
               "Sodium" = sod,
               "Potassium" = pot,
               "Hemoglobin" = hemo,
               "Packed_cell_volume" = pcv,
               "White_blood_cell_count" = wc,
               "Red_blood_cell_count" = rc,
               "Hypertension" = htn,
               "Diabetes_mellitus" = dm,
               "Coronary_artery_disease" = cad,
               "Apetite" = appet,
               "Pedal_edema" = pe,
               "Anemia" = ane,
               "Class" = classification)
```

## Recode the response variable
```{r warning=FALSE, message=FALSE}
data$Class <- ifelse(data$Class == "ckd", 1, 0)
data$Class <- as.factor(data$Class)
```

## Univariate Analysis
## Summary statistics
```{r warning=FALSE, message=FALSE}
data %>% select(2,3,4,5,6,11,12,14,15,16,17,18,19) %>% summary()
```

## Frequency Distribution
## Theme set
```{r warning=FALSE, message=FALSE}
theme_set(theme_classic()+theme(title = element_text
                                (face = "bold", 
                                  colour = "steelblue")))
```

## Age (Histogram)
```{r warning=FALSE, message=FALSE}
p1 <- ggplot(data, aes(x = age)) +
  geom_histogram(fill = "blue", color = "black", bins = 30, alpha = 0.7) +
  labs(title = "Age Distribution", x = "Age", y = "Count") +
  theme_minimal()
```

## Blood Pressure (Histogram)
```{r warning=FALSE, message=FALSE}
p2 <- ggplot(data, aes(x = Blood_pressure)) +
  geom_histogram(fill = "red", color = "black", bins = 12, alpha = 0.7) +
  labs(title = "Blood Pressure Distribution", x = "Blood Pressure", y = "Count") +
  theme_minimal()
```

## Specific Gravity (Histogram)
```{r warning=FALSE, message=FALSE}
p3 <- ggplot(data, aes(x = Specific_graphity)) +
  geom_histogram(fill = "green", color = "black", bins = 5, alpha = 0.7) +
  labs(title = "Specific Gravity Distribution", x = "Specific Gravity", y = "Count") +
  theme_minimal()
```

# Albumin (Histogram)
```{r warning=FALSE, message=FALSE}
p4 <- ggplot(data, aes(x = Albumin)) +
  geom_histogram(fill = "purple", color = "black", bins = 6, alpha = 0.7) +
  labs(title = "Albumin Distribution", x = "Albumin", y = "Count") +
  theme_minimal()
```

## Sugar (Histogram)
```{r warning=FALSE, message=FALSE}
p5 <- ggplot(data, aes(x = Sugar)) +
  geom_histogram(fill = "orange", color = "black", bins = 7, alpha = 0.7) +
  labs(title = "Sugar Distribution", x = "Sugar", y = "Count") +
  theme_minimal()
```

## Red Blood Cells (Bar Plot)
```{r warning=FALSE, message=FALSE}
p6 <- ggplot(data, aes(x = Red_blood_cells, fill = Red_blood_cells)) +
  geom_bar() +
  labs(title = "Red Blood Cells Count", x = "Red Blood Cells", y = "Count") +
  theme_minimal() +
  theme(legend.position = "none")
```

## Pus Cell (Bar Plot)
```{r warning=FALSE, message=FALSE}
p7 <- ggplot(data, aes(x = Pus_cell, fill = Pus_cell)) +
  geom_bar() +
  labs(title = "Pus Cell Count", x = "Pus Cell", y = "Count") +
  theme_minimal() +
  theme(legend.position = "none")
```

## Pus Cell Clumps (Bar Plot)
```{r warning=FALSE, message=FALSE}
p8 <- ggplot(data, aes(x = pus_cell_clamps, fill = pus_cell_clamps)) +
  geom_bar() +
  labs(title = "Pus Cell Clumps", x = "Pus Cell Clumps", y = "Count") +
  theme_minimal() +
  theme(legend.position = "none")
```

## Bacteria (Bar Plot)
```{r warning=FALSE, message=FALSE}
p9 <- ggplot(data, aes(x = Bacteria, fill = Bacteria)) +
  geom_bar() +
  labs(title = "Bacteria Presence", x = "Bacteria", y = "Count") +
  theme_minimal() +
  theme(legend.position = "none")
```

## Blood Glucose Random (Histogram)
```{r warning=FALSE, message=FALSE}
p10 <- ggplot(data, aes(x = Blood_glucose_random)) +
  geom_histogram(fill = "cyan", color = "black", bins = 15, alpha = 0.7) +
  labs(title = "Blood Glucose Random Distribution", x = "Blood Glucose Random", y = "Count") +
  theme_minimal()
```

## Blood Urea (Histogram)
```{r warning=FALSE, message=FALSE}
p11 <- ggplot(data, aes(x = Blood_urea)) +
  geom_histogram(fill = "brown", color = "black", bins = 15, alpha = 0.7) +
  labs(title = "Blood Urea Distribution", x = "Blood Urea", y = "Count") +
  theme_minimal()
```

## Serum Creatinine (Histogram)
```{r warning=FALSE, message=FALSE}
p12 <- ggplot(data, aes(x = Serum_creatanine)) +
  geom_histogram(fill = "pink", color = "black", bins = 30, alpha = 0.7) +
  labs(title = "Serum Creatinine Distribution", x = "Serum Creatinine", y = "Count") +
  theme_minimal()
```

## Sodium (Histogram)
```{r warning=FALSE, message=FALSE}
p13 <- ggplot(data, aes(x = Sodium)) +
  geom_histogram(fill = "yellow", color = "black", bins = 30, alpha = 0.7) +
  labs(title = "Sodium Distribution", x = "Sodium", y = "Count") +
  theme_minimal()
```

## Potassium (Histogram)
```{r warning=FALSE, message=FALSE}
p14 <- ggplot(data, aes(x = Potassium)) +
  geom_histogram(fill = "blue", color = "black", bins = 30, alpha = 0.7) +
  labs(title = "Potassium Distribution", x = "Potassium", y = "Count") +
  theme_minimal()
```

## Hemoglobin (Histogram)
```{r warning=FALSE, message=FALSE}
p15 <- ggplot(data, aes(x = Hemoglobin)) +
  geom_histogram(fill = "green", color = "black", bins = 30, alpha = 0.7) +
  labs(title = "Hemoglobin Distribution", x = "Hemoglobin", y = "Count") +
  theme_minimal()
```

## Hypertension (Bar Plot)
```{r warning=FALSE, message=FALSE}
p16 <- ggplot(data, aes(x = Hypertension, fill = Hypertension)) +
  geom_bar() +
  labs(title = "Hypertension Status", x = "Hypertension", y = "Count") +
  theme_minimal() +
  theme(legend.position = "none")
```

## Diabetes Mellitus (Bar Plot)
```{r warning=FALSE, message=FALSE}
p17 <- ggplot(data, aes(x = Diabetes_mellitus, fill = Diabetes_mellitus)) +
  geom_bar() +
  labs(title = "Diabetes Mellitus Status", x = "Diabetes Mellitus", y = "Count") +
  theme_minimal() +
  theme(legend.position = "none")
```

## Coronary Artery Disease (Bar Plot)
```{r warning=FALSE, message=FALSE}
p18 <- ggplot(data, aes(x = Coronary_artery_disease, fill = Coronary_artery_disease)) +
  geom_bar() +
  labs(title = "Coronary Artery Disease Status", x = "Coronary Artery Disease", y = "Count") +
  theme_minimal() +
  theme(legend.position = "none")
```

## Appetite (Bar Plot)
```{r warning=FALSE, message=FALSE}
p19 <- ggplot(data, aes(x = Apetite, fill = Apetite)) +
  geom_bar() +
  labs(title = "Appetite Status", x = "Appetite", y = "Count") +
  theme_minimal() +
  theme(legend.position = "none")
```

## Pedal Edema (Bar Plot)
```{r warning=FALSE, message=FALSE}
p20 <- ggplot(data, aes(x = Pedal_edema, fill = Pedal_edema)) +
  geom_bar() +
  labs(title = "Pedal Edema Status", x = "Pedal Edema", y = "Count") +
  theme_minimal() +
  theme(legend.position = "none")
```

## Anemia (Bar Plot)
```{r warning=FALSE, message=FALSE}
p21 <- ggplot(data, aes(x = Anemia, fill = Anemia)) +
  geom_bar() +
  labs(title = "Anemia Status", x = "Anemia", y = "Count") +
  theme_minimal() +
  theme(legend.position = "none")
```

## Class (Bar Plot)
```{r warning=FALSE, message=FALSE}
p22 <- ggplot(data, aes(x = Class, fill = Class)) +
  geom_bar() +
  labs(title = "Class Distribution", x = "Class", y = "Count") +
  theme_minimal() +
  theme(legend.position = "none")
```

## Arrange all plots in grids
```{r warning=FALSE, message=FALSE}
grid.arrange(p1, p2, p3, p4, p5, ncol = 2)
```

```{r warning=FALSE, message=FALSE}
grid.arrange(p6, p7, p8, p9, p10, ncol = 2)
```

```{r warning=FALSE, message=FALSE}
grid.arrange(p11, p12, p13, p14, p15, ncol = 2)
```

```{r warning=FALSE, message=FALSE}
grid.arrange(p16, p17, p18, p19, p20, ncol = 2)
```

```{r warning=FALSE, message=FALSE}
grid.arrange(p21, p22, ncol = 2)
```

## Relationship among numeric columns
## Correlation Analysis
```{r warning=FALSE, message=FALSE}
numeric_vars <- data[, sapply(data, is.numeric)]  # Select only numeric variables
cor_matrix <- cor(numeric_vars, use = "complete.obs")
ggcorrplot(cor_matrix,title = "Correlation Matrix",lab_col = "black",
           lab = TRUE, legend.title = "Pearson Correlation",
           lab_size = 3, ggtheme = theme_classic(),
           outline.color = "black",
           colors = c("orange", "green", "blue"))

```

## Relationship between Class and other categorical variables
## Define categorical variables
```{r warning=FALSE, message=FALSE}
categorical_vars <- c("Red_blood_cells", "Pus_cell", "pus_cell_clamps", "Bacteria",
                      "Hypertension", "Diabetes_mellitus", "Coronary_artery_disease",
                      "Apetite", "Pedal_edema", "Anemia")
```

## Loop through each categorical variable
```{r warning=FALSE, message=FALSE}
for (var in categorical_vars) {
  cat("Chi-square test between Class and", var, "\n")
  chisq_test <- chisq.test(table(data[[var]], data$Class))
  print(chisq_test)
  
  # Bar chart visualization
  plot <- ggplot(data, aes(x = .data[[var]], fill = Class)) +
    geom_bar(position = "dodge") +
    labs(title = paste("Class vs", var), x = var, y = "Count") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_fill_brewer(palette = "Set1")
  
  print(plot)  # Display the plot
}
```

## Relationshi between Class and other numeric variables
## Define numeric variables
```{r warning=FALSE, message=FALSE}
numeric_vars <- c("age", "Blood_pressure", "Specific_graphity", "Albumin",
                  "Sugar", "Blood_glucose_random", "Blood_urea", "Serum_creatanine",
                  "Sodium", "Potassium", "Hemoglobin", "Red_blood_cell_count")
```

## Loop through each numeric variable
```{r warning=FALSE, message=FALSE}
for (var in numeric_vars) {
  cat("T-test between Class and", var, "\n")
  t_test <- t.test(data[[var]] ~ data$Class, var.equal = TRUE)
  print(t_test)
  
 ## Boxplot visualization
  plot <- ggplot(data, aes(x = Class, y = .data[[var]], fill = Class)) +
    geom_boxplot() +
    labs(title = paste("Class vs", var), x = "Class", y = var) +
    theme_minimal() +
    scale_fill_brewer(palette = "Set2")
  
  print(plot)  # Display the plot
}
```

## Multivariate analysis
```{r warning=FALSE, message=FALSE}
library(glmnet)
x <- model.matrix(Class ~ ., data)[, -1]  # Remove intercept
y <- data$Class
```

# Fit Lasso (alpha = 1) or Ridge (alpha = 0)
```{r warning=FALSE, message=FALSE}
cv_model <- cv.glmnet(x, y, family = "binomial", alpha = 1)  # Lasso regression
best_lambda <- cv_model$lambda.min  # Get optimal lambda
```

# Refit using the best lambda
```{r warning=FALSE, message=FALSE}
final_model <- glmnet(x, y, family = "binomial", alpha = 1, lambda = best_lambda)
print(final_model)
```

## Interpret the Model Coefficients
```{r warning=FALSE, message=FALSE}
coef(final_model)
```

## Make predictions
```{r warning=FALSE, message=FALSE}
pred_prob <- predict(final_model, newx = x, type = "response")  
head(pred_prob)  # Show first few predicted probabilities  
pred_class <- ifelse(pred_prob > 0.5, 1, 0)  
head(pred_class)  # Show first few predicted classes  
```

## Evaluate Model Performance
```{r warning=FALSE, message=FALSE}
confusionMatrix(as.factor(pred_class), as.factor(y))
```

## Produce ROC curve
```{r warning=FALSE, message=FALSE}
library(pROC)
roc_curve <- roc(y, as.numeric(pred_prob))
plot(roc_curve, col = "blue", main = "ROC Curve for Lasso Model")
auc(roc_curve)  # Get AUC value
```

