---
title: "CLINICAL STUDY OF PROSTATE CANCER"
author: "Enock Bereka"
date: "2025-02-10"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Load necessary libraries
```{r warning=FALSE, message=FALSE}
library(naniar)
library(tidyverse)
library(pROC)
library(performance)
library(ggeffects)
library(sjPlot)
library(vip)
library(report)
library(car)
library(gtsummary)
library(ggcorrplot)
```

## Load and inspect the dataset
```{r warning=FALSE, message=FALSE}
data <- cardata <- read_csv("C:/Users/PC/OneDrive/Desktop/Data Science/Datasets/Datasets/Prostate_Cancer.csv")
glimpse(data)
vis_miss(data)
sum(duplicated(data))
unique(data$diagnosis_result)
```

## Data Preparation
```{r warning=FALSE, message=FALSE}
data$diagnosis_result <- ifelse(data$diagnosis_result == "M", 1, 0)
data$diagnosis_result <- as.factor(data$diagnosis_result)
```

## Univariate analysis
## Summary statistics
```{r warning=FALSE, message=FALSE}
data %>% select(-1,-2) %>% summary()

## Frequency distributions for numeric variables
## Theme set
theme_set(theme_minimal()+theme(title = element_text
                                (face = "bold", 
                                  colour = "steelblue")))
```

```{r warning=FALSE, message=FALSE}
## Radius
ggplot(data, aes(x = radius))+
  geom_density(fill = "purple", color = "red")+
  ggtitle("Distribution of Subjects based on Radius")
```

## Texture
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(x = texture))+
  geom_density(fill = "purple", color = "red")+
  ggtitle("Distribution of Subjects based on Texture")
```


## Perimeter
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(x = perimeter))+
  geom_density(fill = "purple", color = "red")+
  ggtitle("Distribution of Subjects based on Perimeter")
```

## Area
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(x = area))+
  geom_density(fill = "purple", color = "red")+
  ggtitle("Distribution of Subjects based on Area")
```

## Smoothness
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(x = smoothness))+
  geom_density(fill = "purple", color = "red")+
  ggtitle("Distribution of Subjects based on Smoothness")
```

## Compactness
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(x = compactness))+
  geom_density(fill = "purple", color = "red")+
  ggtitle("Distribution of Subjects based on Compactness")
```

## Symmetry
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(x = symmetry))+
  geom_density(fill = "purple", color = "red")+
  ggtitle("Distribution of Subjects based on Symetry")
```

## Fractal Dimension
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(x = fractal_dimension))+
  geom_density(fill = "purple", color = "red")+
  ggtitle("Distribution of Subjects based on Fractal Dimension")
```

## Distribution of the outcome variabe
```{r warning=FALSE, message=FALSE}
table(data$diagnosis_result)
ggplot(data, aes(diagnosis_result, fill = diagnosis_result))+
  geom_bar()+
  ggtitle("Distribution of Subjects based on diagnosis results")+
  scale_fill_discrete(name = "diagnosis_result", label = c("Negative", "Positive"))
glimpse(data)
```

## There were 62 subjects with prostate cancer(cases) and 38 subjects without prostate cancer(controls)

## Bivariate analysis
## Correlation Analysis
```{r warning=FALSE, message=FALSE}
cr <- data %>% select(-1,-2) %>% cor()
ggcorrplot(cr,title = "Correlation Matrix",lab_col = "black",
           lab = TRUE, legend.title = "Pearson Correlation",
           lab_size = 3, ggtheme = theme_classic(),
           outline.color = "black",
           colors = c("orange", "green", "blue"))
```

## There is a high positive correlation between area and perimeter
## There is a slighly low positive correlation among the remaining variables.

##T-Test
## Radius vs Diagnostic results
```{r warning=FALSE, message=FALSE}
t.test(radius~diagnosis_result, data = data)
```

## There is no statistically significant difference in mean of radius of a tumor between positive and negative cases indicating minimal association between the variables.

## Visualize results
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(x = diagnosis_result, y = radius,
                 fill = diagnosis_result))+
  geom_boxplot()+
  ggtitle("Radius vs Diagnostic results")+
  scale_fill_discrete(name = "diagnosis_result", label = c("Negative", "Positive"))
```

## Texture vs Diagnostic results
```{r warning=FALSE, message=FALSE}
t.test(texture~diagnosis_result, data = data)
```

## There is no statistically significant difference in mean of textuer of a tumor between positive and negative cases indicating minimal association between the variables.


## Visualize results
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(x = diagnosis_result, y = texture,
                 fill = diagnosis_result))+
  geom_boxplot()+
  ggtitle("Texture vs Diagnostic results")+
  scale_fill_discrete(name = "diagnosis_result", label = c("Negative", "Positive"))
```

## Perimeter vs Diagnostic results
```{r warning=FALSE, message=FALSE}
t.test(perimeter~diagnosis_result, data = data)
```

## There is a statistically significant difference in mean of perimeter of a tumor between positive and negative cases indicating strong association between the variables.

## Visualize results
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(x = diagnosis_result, y = perimeter,
                 fill = diagnosis_result))+
  geom_boxplot()+
  ggtitle("Perimeter vs Diagnostic results")+
  scale_fill_discrete(name = "diagnosis_result", label = c("Negative", "Positive"))
```


## Compactness vs Diagnostic results
```{r warning=FALSE, message=FALSE}
t.test(compactness~diagnosis_result, data = data)
```

## There is a statistically significant difference in mean of compactness of a tumor between positive and negative cases indicating strong association between the variables.

## Visualize results
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(x = diagnosis_result, y = compactness,
                 fill = diagnosis_result))+
  geom_boxplot()+
  ggtitle("Compactness vs Diagnostic results")+
  scale_fill_discrete(name = "diagnosis_result", label = c("Negative", "Positive"))
```

## Symmetry vs Diagnostic results
```{r warning=FALSE, message=FALSE}
t.test(symmetry~diagnosis_result, data = data)
```

## There is a statistically significant difference in mean of symmetry of a tumor between positive and negative cases indicating strong association between the variables.

## Visualize results
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(x = diagnosis_result, y = symmetry,
                 fill = diagnosis_result))+
  geom_boxplot()+
  ggtitle("Symmetry vs Diagnostic results")+
  scale_fill_discrete(name = "diagnosis_result", label = c("Negative", "Positive"))
```


## Fractal Dimension vs Diagnostic results
```{r warning=FALSE, message=FALSE}
t.test(fractal_dimension~diagnosis_result, data = data)
```

## There is no statistically significant difference in mean of fractal dimension of a tumor between positive and negative cases indicating minimal association between the variables.

## Visualize results
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(x = diagnosis_result, y = fractal_dimension,
                 fill = diagnosis_result))+
  geom_boxplot()+
  ggtitle("Fractal dimension vs Diagnostic results")+
  scale_fill_discrete(name = "diagnosis_result", label = c("Negative", "Positive"))
```

## Multivariate Analysis
```{r warning=FALSE, message=FALSE}
data$id <- NULL
model <- glm(diagnosis_result~., data = data, family = binomial)
```

## Check model assumptions
## Goodness of fit
```{r warning=FALSE, message=FALSE}
check_predictions(model) %>% plot()
```

## Homoscedasticity
```{r warning=FALSE, message=FALSE}
check_residuals(model)
check_residuals(model) %>% plot()
```

## Outliers
```{r warning=FALSE, message=FALSE}
check_outliers(model)
check_outliers(model) %>% plot()
```

## Autocorrelation
```{r warning=FALSE, message=FALSE}
check_autocorrelation(model)
```

## Check the models predicted probabilities
```{r warning=FALSE, message=FALSE}
ggeffect(model)
```

## Visualize the models predicted probabilities
```{r warning=FALSE, message=FALSE}
ggeffect(model) %>% plot() %>% sjPlot::plot_grid()
```

## An increase in radius of a tumor leads to an incease in risk of developing prostate cancer.
## Higher texture values are associated with malignant tumors.
## High values of perimeter of tumors are associated with malignant turmors.
## An increase in area of a turmor leads to development of prostate cancer cell.
## An increase in smoothness of a tumor leads to a decrease in development of prostate cancer cell.
## High compactness of tumors is associated with malignancy.
## Lower symmetry of tumors is associted with malignancy.
## Higher fractal dimension is associted with malignancy.

## Check for odds ratios
```{r warning=FALSE, message=FALSE}
fancy_table <- tbl_regression(
  model, exponentiate = TRUE,
  add_pairwise_contrasts = TRUE,
  contrasts_adjust = "bonferroni",
  pairwise_reverse = TRUE,
  pvalue_fun = ~style_pvalue(.x, digits = 3)) %>% 
  add_significance_stars(hide_p = F, hide_se = T,
                         hide_ci = F) %>% 
  bold_p()
fancy_table
```

## Feature importance plot
```{r warning=FALSE, message=FALSE}
car::Anova(model)
vip(model)
```

## Performance of the model
```{r warning=FALSE, message=FALSE}
performance(model)
```

## The model explains a substantial variance of 57.7 percent in the outcome variable indicating a better fit.
## How good is our model in classification
```{r warning=FALSE, message=FALSE}
roc(diagnosis_result~fitted.values(model), data = data,
    plot = TRUE, legacy.axes = TRUE, 
    print.auc = TRUE, ci = TRUE)
```

## The model achieves an AUC of 0.929 making it a valuable tool for classification.