---
title: "Diabetes Analysis Project"
author: "Enock Bereka"
date: "2025-01-25"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load necessary libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(reshape2)
library(ggfortify)
library(skimr)
library(kableExtra)
library(patchwork)
library(directlabels)
library(MASS)
library(hrbrthemes)
library(viridis)
library(RColorBrewer)
library(ggcorrplot)
```

## Load and inspect the dataset
```{r warning=FALSE, message=FALSE}
diabetes <- read_csv("C:/Users/PC/OneDrive/Desktop/Data Science/Datasets/Datasets/diabetes.csv")
glimpse(diabetes)
anyNA(diabetes)
duplicated(diabetes) %>% table()
```

## Data preparation
## Assign ID to act as a unique identifier
```{r}
diabetes$id <- seq.int(nrow(diabetes))
```

## Convert response variable to binary
```{r}
diabetes$outcome <- ifelse(diabetes$Outcome==0, "Negative", "Positive")
```

## Variable categorisation.
```{r}
diabetes$pregnancy <- ifelse(diabetes$Pregnancies<=8, "Under 8", "Above 8")
diabetes$Glucose_level <- ifelse(diabetes$Glucose>=70 & diabetes$Glucose<=100, "Glucose Normal", "Glucose Abnormal")
diabetes$blood_pressure <- ifelse(diabetes$BloodPressure>=90 & diabetes$BloodPressure<=120, "Normal", "Abnormal")
diabetes$bmi <- ifelse(diabetes$BMI<18.5, "Underweight",                    ifelse(diabetes$BMI>=18.5&diabetes$BMI<25, "Normal Weight", ifelse(diabetes$BMI>=25&diabetes$BMI<30, "Overweight", ifelse(diabetes$BMI>=30&diabetes$BMI<40, "Obese", "Severely Obese"))))
diabetes$dpf <- ifelse(diabetes$DiabetesPedigreeFunction <= 0.70, "Normal", "Abnormal")
diabetes$age <- ifelse(diabetes$Age>=21&diabetes$Age<=35, "Young",                     ifelse(diabetes$Age>=36&diabetes$Age<=55, "Middle Aged", "Old"))
diabetes$Skinthickness <- ifelse(diabetes$SkinThickness<=23, "Normal", "Abnormal")
diabetes$insulin <- ifelse(diabetes$Insulin<=25, "Insulin_Normal", "Insulin_Abnormal")
```

## Original Data table
```{r}
diabetes1 <- subset(diabetes, select=c(Pregnancies, Glucose, BloodPressure, SkinThickness, Insulin, BMI,DiabetesPedigreeFunction,Age,Outcome))
```

## Univariate Analysis
## Summary statistics for numeric columns
```{r}
diabetes2 <- subset(diabetes, select=c(Pregnancies, Glucose, BloodPressure, SkinThickness, Insulin, BMI,DiabetesPedigreeFunction,Age))
summary(diabetes2)
```

## Binary Variable Distribution 
```{r}
#1. Age
p1 <-ggplot(diabetes, aes(x=age))+geom_bar(fill="green", color = "black")+labs(x="Age Group", title = "Age Distribution")+theme_minimal(base_size=10)
#2. Pregnancies
p2 <-ggplot(diabetes, aes(x=pregnancy))+geom_bar(fill="green", color = "black")+labs(x="Pregnancies", title = "Number of pregnancies distribution")+theme_minimal(base_size=10)
#3. Glucose Levels
p3 <-ggplot(diabetes, aes(x=Glucose_level))+geom_bar(fill="green", color = "black")+labs(x="Glucose Level", title = "Distribution of glucose levels")+theme_minimal(base_size=10)
#4. Insulin
p4 <-ggplot(diabetes, aes(x=insulin))+geom_bar(fill="green", color = "black")+labs(x="Insulin Levels", title = "Insulin Level distribution")+theme_minimal(base_size=10)
#5. Blood Pressure
p5 <-ggplot(diabetes, aes(x=blood_pressure))+geom_bar(fill="green", color = "black")+labs(x="Blood Pressure", title = "Blood Pressure distribution")+theme_minimal(base_size=10)
#6. BMI
p6 <-ggplot(diabetes, aes(x=bmi))+geom_bar(fill="green", color = "black")+labs(x="BMI Levels", title = "BMI Distribution")+theme_minimal(base_size=10)
#7. Skin Thickness
p7 <-ggplot(diabetes, aes(x=Skinthickness))+geom_bar(fill="green", color = "black")+labs(x="Skin Thickness", title = "Skin Thickness distribution")+theme_minimal(base_size=10)
#8. Diabetes Pedigree Function
p8 <-ggplot(diabetes, aes(x=dpf))+geom_bar(fill="green", color = "black")+labs(x="Diabetes Pedigree Function", title = "Diabetes Pedigree Function Distribution")+theme_minimal(base_size=10)
#8. Diabetes Outcome
p9 <-ggplot(diabetes, aes(x=outcome))+geom_bar(fill="green", color = "black")+labs(x="Outcome", title = "Prevalence of Diabetes")+theme_minimal(base_size=10)
```

## Demographic Distribution
```{r}
(p1+p2)+
  plot_annotation(title="Demographic Distribution")
```

## Biochemical Factors Distribution
```{r}
(p3+p4)+
  plot_annotation(title = "Biochemical Factors Distribution")
```

## Physical Factors Distribution.
```{r}
(p5+p6+p7)+
  plot_annotation(title = "Physical Factors Distribution")
```

## Lab Test Distribution
```{r}
(p8+p9)+
  plot_annotation(title = "Lab Test Distribution")
```

## Continuous Variables Disbribution
#1. Age
```{r}
ggplot(diabetes1, aes(x=Age, fill = ..count..))+
  scale_fill_discrete("Frequency", low = "yellow", high = "red")+
  geom_histogram(binwidth=5, alpha=0.8)+
  geom_vline(xintercept = 35, linetype="dashed")+
  geom_vline(xintercept = 55, linetype="dashed")+
  annotate("text", x=25, y=100, label="Young", size=2.5, color="red") +
  annotate("text", x=47, y=100, label="Middle Age", size=2.5, color="red") +
  annotate("text", x=70, y=100, label="Old", size=2.5, color="red") +
  labs(title="Age Distribution") + theme_minimal(base_size = 8)
```

#2. Pregnancies
```{r}
ggplot(diabetes1, aes(x=Pregnancies, fill = ..count..))+ geom_histogram(binwidth=5, alpha=0.8)+
scale_fill_discrete("Frequency", low = "yellow", high = "red")+
  geom_vline(xintercept = 8, linetype="dashed")+
  annotate("text", x=4, y=400, label="Low Risk", size=2.5, color="red") +
  annotate("text", x=16, y=400, label="High Risk", size=2.5, color="red") +
  labs(title="Pregnancies Distribution") + theme_minimal(base_size = 8)
```

#3. Glucose Levels
```{r}
ggplot(diabetes1, aes(x=Glucose, fill = ..count..))+ geom_histogram(binwidth=5,  alpha=0.8)+
 scale_fill_discrete("Frequency", low = "yellow", high = "red")+
  geom_vline(xintercept = 70, linetype="dashed")+
  annotate("text", x=35, y=50, label="Normal", size=2.5, color="red") +
  annotate("text", x=160, y=50, label="Abnormal", size=2.5, color="red") +
  labs(title="Glucose Distribution") + theme_minimal(base_size = 8)
```

#4. Insulin
```{r}
ggplot(diabetes1, aes(x=Insulin, fill = ..count..))+ geom_histogram(binwidth=5, alpha=0.8)+
  scale_fill_discrete("Frequency", low = "yellow", high = "red")+
  geom_vline(xintercept = 25, linetype="dashed")+
  annotate("text", x=15, y=50, label="Normal", size=2.5, color="red") +
  annotate("text", x=100, y=50, label="Abnormal", size=2.5, color="red") +
  labs(title="Insulin Distribution") + theme_minimal(base_size = 8)
```

#5. Blood Pressure
```{r}
ggplot(diabetes1, aes(x=BloodPressure, fill = ..count..))+ geom_histogram(binwidth=5, alpha=0.8)+
  scale_fill_discrete("Frequency", low = "yellow", high = "red")+
  geom_vline(xintercept = 90, linetype="dashed")+
  geom_vline(xintercept = 120, linetype="dashed")+
  annotate("text", x=100, y=50, label="Normal", size=2.5, color="red") +
  annotate("text", x=50, y=50, label="Abnormal", size=2.5, color="red") +
  labs(title= "Blood Pressure Distribution") + theme_minimal(base_size = 8)
```

#6. BMI
```{r}
ggplot(diabetes1, aes(x=BMI, fill = ..count..))+ geom_histogram(binwidth=5, alpha=0.8)+
  scale_fill_discrete("Frequency", low = "yellow", high = "red")+
  geom_vline(xintercept = 18.5, linetype="dashed")+
  geom_vline(xintercept = 25, linetype="dashed")+
  geom_vline(xintercept = 30, linetype="dashed")+
  geom_vline(xintercept = 40, linetype="dashed")+
  annotate("text", x=14, y=50, label="Underweight", size=2.5, color="red") +
  annotate("text", x=22, y=60, label="Normalweight", size=2.5, color="red") +
  annotate("text", x=27, y=70, label="Overweight", size=2.5, color="red") +
  annotate("text", x=35, y=80, label="Obese", size=2.5, color="red") +
  annotate("text", x=45, y=50, label="Severelyobese", size=2.5, color="red") +
  labs(title= "BMI Distribution") + theme_minimal(base_size = 8)
```

#7. Skin Thickness
```{r}
ggplot(diabetes1, aes(x=SkinThickness, fill = ..count..))+ geom_histogram(binwidth=5, alpha=0.8)+
  scale_fill_discrete("Frequency", low = "yellow", high = "red")+
  geom_vline(xintercept = 23, linetype="dashed")+
  annotate("text", x=15, y=50, label="Normal", size=2.5, color="red") +
  annotate("text", x=30, y=50, label="Abnormal", size=2.5, color="red") +
  labs(title= "Skin Thickness Distribution") + theme_minimal(base_size = 8)
```

## Bivariate Analysis
## Relationship between continous variables
## Correlation
```{r}
cr <- cor(diabetes2)
ggcorrplot(cr,title = "Correlation Heatmap",lab_col = "black",
           lab = TRUE, legend.title = "Pearson Correlation",
           lab_size = 3, ggtheme = theme_classic(),
           outline.color = "black",
           colors = c("orange", "green", "blue"))
```

## T-test
## Relationship between Age and Diabetes Outcome
```{r}
t.test(Age~outcome, data = diabetes)
```

## Relationship between Insulin and Diabetes Outcome
```{r}
t.test(Insulin~outcome, data = diabetes)
```

## Relationship between BMI and Diabetes Outcome
```{r}
t.test(BMI~outcome, data = diabetes)
```

## Relationship between Glucose and Diabetes Outcome
```{r}
t.test(Glucose~outcome, data = diabetes)
```

## Relationship between Age and BMI levels
```{r}
a <- aov(Age~bmi, data = diabetes)
summary(a)
```

## Relationship between Glucose and BMI levels
```{r}
a <- aov(Glucose~bmi, data = diabetes)
summary(a)
```

## Relationship between Insulin and Age groups
```{r}
a <- aov(Insulin~age, data = diabetes)
summary(a)
```

## Multivariate analysis
## Logistic Regression model
```{r}
model <- glm(Outcome~., data = diabetes1, family = binomial)
```

## Check Model assumptions
```{r warning=FALSE, message=FALSE}
library(performance)
check_autocorrelation(model)
check_collinearity(model)
check_predictions(model) %>% plot()
check_residuals(model)
check_residuals(model) %>% plot()
check_outliers(model)
check_outliers(model) %>% plot()
```

## Get the summary of the model
```{r}
summary(model)
```

## Check model probabilities
```{r warning=FALSE, message=FALSE}
library(ggeffects)
ggeffect(model)
```

## Visualize model predictions
```{r warning=FALSE, message=FALSE}
library(sjPlot)
fancy_plot <- ggpredict(model) %>% plot() %>% 
  sjPlot::plot_grid()
```

## Check if model predictions are statistically significant
```{r warning=FALSE, message=FALSE}
library(gtsummary)
fancy_table <- tbl_regression(
  model, exponentiate = TRUE,
  add_pairwise_contrasts = TRUE,
  contrasts_adjust = "bonferroni",
  pairwise_reverse = FALSE,
  pvalue_fun = ~style_pvalue(.x, digits = 3)) %>% 
  add_significance_stars(hide_p = F, hide_se = T,
                         hide_ci = F) %>% 
  bold_p()
fancy_table
```

## Checking variable importance
```{r warning=FALSE, message=FALSE}
library(vip)
car::Anova(model)
vip(model)
```

## How accurate is our model in making predictions
## Produce ROC curve
```{r warning=FALSE, message=FALSE}
library(pROC)
roc(Outcome~fitted.values(model), data = diabetes,
    plot = TRUE, legacy.axes = TRUE, 
    print.auc = TRUE, ci = TRUE)
```

