---
title: "HEALTHCARE ANALYSIS"
author: "Enock Bereka"
date: "2025-02-06"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load necessary libraries
```{r warning=FALSE, message=FALSE}
library(tidyverse)    # For data manipulation and visualization
library(ggplot2)      # For visualizations
library(dplyr)        # For data wrangling
library(ggcorrplot)     # For correlation analysis
library(caret)        # For machine learning
library(lubridate)    # For date handling
library(ggpubr)       # For statistical analysis
library(cluster)      # For clustering
library(randomForest) # For machine learning
library(forecast)     # For time series analysis
```

## Load and inspect the dataset
```{r warning=FALSE, message=FALSE}
healthcare_data <- read_csv("C:/Users/PC/Downloads/healthcare_dataset.csv")
glimpse(healthcare_data)
anyNA(healthcare_data)
duplicated(healthcare_data) %>% table()
```

## Convert date columns to Date format
```{r warning=FALSE, message=FALSE}
healthcare_data$`Date of Admission` <- as.Date(healthcare_data$`Date of Admission`, format="%Y-%m-%d")
healthcare_data$`Discharge Date` <- as.Date(healthcare_data$`Discharge Date`, format="%Y-%m-%d")
```

# Handle missing values by replacing them with the median (for numeric) or mode (for categorical)
```{r warning=FALSE, message=FALSE}
healthcare_data <- healthcare_data %>%
  mutate(across(where(is.numeric), ~replace_na(., median(., na.rm=TRUE)))) %>%
  mutate(across(where(is.factor), ~replace_na(., as.character(mode(na.omit(.))))))
```

## Remove duplicate rows
```{r warning=FALSE, message=FALSE}
healthcare_data <- distinct(healthcare_data)
```

## Univariate Analysis
## Summary Statistics for Age, Billing Amound And Room Numbers
```{r warning=FALSE, message=FALSE}
healthcare_data %>% select(Age, `Billing Amount`, `Room Number`) %>% summary()
```

## Majority of the patients were aroung 51 years
## Malority of the patient incurred around 25554 on medical costs
## Majority of the patients were in room number 301
## What is the distribution of patient ages?
```{r warning=FALSE, message=FALSE}
ggplot(healthcare_data, aes(x=Age)) + 
  geom_histogram(fill="green",color = "black", bins=15, alpha=0.7) +
  theme_minimal() +
  ggtitle("Age Distribution of Patients")
```

## What is the distribution of Patients based on Medical bills?
```{r warning=FALSE, message=FALSE}
ggplot(healthcare_data, aes(x=`Billing Amount`)) +
geom_histogram(fill="purple",color = "red", bins=15, alpha=0.7) +
  theme_minimal() +
  ggtitle("Distribution of Hospital Billing Amounts")
```

## What is the distribution of Patients based on Room Number?
```{r warning=FALSE, message=FALSE}
ggplot(healthcare_data, aes(x=`Room Number`)) +
  geom_histogram(fill="purple",color = "red", bins=15, alpha=0.7) +
  theme_minimal() +
  ggtitle("Distribution of Patients based on Room Number")
```

## What is the distribution of patients based on common medical conditions?
```{r warning=FALSE, message=FALSE}
table(healthcare_data$`Medical Condition`)
ggplot(healthcare_data, aes(x=`Medical Condition`)) +
  geom_bar(fill="green",color = "black", alpha=0.7) +
  theme_minimal() +
  ggtitle("Distribution of patients based on common medical conditions") +
  coord_flip()
```

## Almost all the medical conditions had similar number of patients

## What is the distribution of patients based on Gender?
```{r warning=FALSE, message=FALSE}
table(healthcare_data$Gender)
ggplot(healthcare_data, aes(x=Gender)) +
  geom_bar(fill="green",color = "black", alpha=0.7) +
  theme_minimal() +
  ggtitle("Distribution of patients based on Gender")
```

## Male patients were slightly higher compared to female patients

## What is the distribution of patients based on Type of admission?
```{r warning=FALSE, message=FALSE}
table(healthcare_data$`Admission Type`)
ggplot(healthcare_data, aes(x=`Admission Type`)) +
  geom_bar(fill="green",color = "black", alpha=0.7) +
  theme_minimal() +
  ggtitle("Distribution of patients based on Type of admission")
```

## Majority of the patients used the Elective admission type.

## What is the distribution of patients based on Blood type?
```{r warning=FALSE, message=FALSE}
table(healthcare_data$`Blood Type`)
ggplot(healthcare_data, aes(x=`Blood Type`)) +
  geom_bar(fill="green",color = "black", alpha=0.7) +
  theme_minimal() +
  ggtitle("Distribution of patients based on Blood type")
```

## Majority of the patients had A- blood type

## What is the distribution of patients based on Insurance provider?
```{r warning=FALSE, message=FALSE}
table(healthcare_data$`Insurance Provider`)
ggplot(healthcare_data, aes(x=`Insurance Provider`)) +
  geom_bar(fill="green",color = "black", alpha=0.7) +
  theme_minimal() +
  ggtitle("Distribution of patients based on Insurance provider")
```

## Majority of the patients used Cigma insurance provider

## What is the distribution of patients based on type of Medication?
```{r warning=FALSE, message=FALSE}
table(healthcare_data$Medication)
ggplot(healthcare_data, aes(x=Medication)) +
  geom_bar(fill="green",color = "black", alpha=0.7) +
  theme_minimal() +
  ggtitle("Distribution of patients based on type of Medication")
```

## Majority of the patients were treated using Lipitor drug

## What is the distribution of patients based on Test Results?
```{r warning=FALSE, message=FALSE}
table(healthcare_data$`Test Results`)
ggplot(healthcare_data, aes(x=`Test Results`)) +
  geom_bar(fill="green",color = "black", alpha=0.7) +
  theme_minimal() +
  ggtitle("Distribution of patients based on Test Results")
```

## The test results for majority of patients was abnormal.

## BIVARIATE ANALYSIS
## Relationship between Age ,Billing Amount and Room Number
## Correlation Analysis 
```{r warning=FALSE, message=FALSE}
num_data <- healthcare_data %>% select_if(is.numeric)
cor_matrix <- cor(num_data)
ggcorrplot(cor_matrix,title = "Correlation Matrix",lab_col = "black",
           lab = TRUE, legend.title = "Pearson Correlation",
           lab_size = 3, ggtheme = theme_classic(),
           outline.color = "black",
           colors = c("orange", "green", "blue"))
```

## There is no relationship between Age, Billing Amount and Room Numbers

## ANOVA: Does billing amount vary by medical condition?
```{r warning=FALSE, message=FALSE}
anova_result <- aov(`Billing Amount` ~ `Medical Condition`, data=healthcare_data)
summary(anova_result)
```

## There is a statistically significant difference in Billing Amount across medical conditions indicating an association between the two.

## Visualize the Anova results
```{r warning=FALSE, message=FALSE}
ggplot(healthcare_data, aes(x=`Medical Condition`, y=`Billing Amount`)) +
  geom_boxplot(fill="orange", alpha=0.7) +
  theme_minimal() +
  ggtitle("Billing Amount per Medical Condition") +
  coord_flip()
```

## Does the length of hospital stay differ based on the type of admission?
## Calculate Length of Stay
```{r warning=FALSE, message=FALSE}
healthcare_data$Length_of_Stay <- as.numeric(healthcare_data$`Discharge Date` - healthcare_data$`Date of Admission`)

## Perform one-way ANOVA
anova_result2 <- aov(Length_of_Stay ~ `Admission Type`, data=healthcare_data)
summary(anova_result2)
```

## Visualize the results
```{r warning=FALSE, message=FALSE}
ggplot(healthcare_data, aes(x=`Admission Type`, y=Length_of_Stay, fill=`Admission Type`)) +
  geom_boxplot() +
  theme_minimal() +
  ggtitle("Length of Hospital Stay by Admission Type")
```

## There is no statistically significant difference in length of stay based on admission types indicating no association between the two.

## Does the test result (normal, abnormal and Inconclusive) affect the billing amount?
## Perform one-way ANOVA: Billing Amount ~ Test Results
```{r warning=FALSE, message=FALSE}
anova_result3 <- aov(`Billing Amount` ~ `Test Results`, data=healthcare_data)
summary(anova_result3)
```

## Boxplot visualization
```{r warning=FALSE, message=FALSE}
ggplot(healthcare_data, aes(x=`Test Results`, y=`Billing Amount`, fill=`Test Results`)) +
  geom_boxplot() +
  theme_minimal() +
  ggtitle("Billing Amount by Test Results")
```

## There is no statistically significant difference in Billing Amount across test results indicating no association.

## Does the average patient age differ across different insurance providers?
# Perform one-way ANOVA
```{r warning=FALSE, message=FALSE}
anova_result4 <- aov(Age ~ `Insurance Provider`, data=healthcare_data)
summary(anova_result4)
```

## There is no statistically significant difference in Age across the insurance providers indicating no association.


## Visualize the distribution
```{r warning=FALSE, message=FALSE}
ggplot(healthcare_data, aes(x=`Insurance Provider`, y=Age, fill=`Insurance Provider`)) +
  geom_boxplot() +
  theme_minimal() +
  ggtitle("Patient Age by Insurance Provider") +
  coord_flip()
```

## T-test: Do men and women have different billing amounts?
```{r warning=FALSE, message=FALSE}
t_test_result <- t.test(`Billing Amount` ~ Gender, data=healthcare_data)
print(t_test_result)
```

## There is no statistically significant difference in Billing Amount between Male and Female indicating no association.

## Chi-square test: Is Admission Type related to Gender?
```{r warning=FALSE, message=FALSE}
chi_sq <- chisq.test(table(healthcare_data$Gender, healthcare_data$`Admission Type`))
print(chi_sq)
```

## There is a statistically significant association between Gender and admission type indicated by small significant p-value and higher Chi-square statistic.

## Visualize the result
```{r warning=FALSE, message=FALSE}
ggplot(healthcare_data, aes(`Admission Type`, fill = Gender))+
  geom_bar(position = "dodge")+
  theme_minimal() +
  ggtitle("Admission Type vs Gender")
```

## TIME SERIES ANALYSIS (Forecasting Admissions)
## Aggregate admissions by month
```{r warning=FALSE, message=FALSE}
admissions_ts <- healthcare_data %>%
  mutate(Admission.Month = floor_date(`Date of Admission`, "month")) %>%
  group_by(Admission.Month) %>%
  summarize(Admissions = n())

## Convert to time series object
ts_data <- ts(admissions_ts$Admissions, start=c(year(min(admissions_ts$Admission.Month))), frequency=12)
```

## General trend
```{r warning=FALSE, message=FALSE}
plot(aggregate(ts_data, FUN = mean), ylab = "Admissions", main = "Monthly Hospital Admissions Over Time")
```

## There is seasonal fluctuation in monthly admissions of patients.

## Plot time series
```{r warning=FALSE, message=FALSE}
autoplot(ts_data) + ggtitle("Monthly Hospital Admissions Over Time")
```

## Plot a season plot
```{r warning=FALSE, message=FALSE}
ggseasonplot(ts_data, main = "Monthly Hospital Admissions Over Time")
```

## Forecast next 12 months using ARIMA
```{r warning=FALSE, message=FALSE}
arima_model <- auto.arima(ts_data)
forecasted_values <- forecast(arima_model, h=12)
forecasted_values
```

## Visualize the forecasts
```{r warning=FALSE, message=FALSE}
autoplot(forecasted_values) + ggtitle("Forecasted Admissions for Next 12 Months")
```

## The forecast indicates that there will be a constant number of admissions in the next 12 months

## COST ANALYSIS
# Top 5 most expensive medical conditions
```{r warning=FALSE, message=FALSE}
top_conditions <- healthcare_data %>%
  group_by(`Medical Condition`) %>%
  summarize(Average_Cost = mean(`Billing Amount`, na.rm=TRUE)) %>%
  arrange(desc(Average_Cost)) %>%
  head(5)
print(top_conditions)
```

## Obesity was the medical condition with the highest medical cost followed by Diabetes, Asthma, Arthritis and then Hypertension

## Average Billing Amount by Insurance Provider
```{r warning=FALSE, message=FALSE}
insurance_costs <- healthcare_data %>%
  group_by(`Insurance Provider`) %>%
  summarize(Average_Billing = mean(`Billing Amount`, na.rm=TRUE))

ggplot(insurance_costs, aes(x=reorder(`Insurance Provider`, Average_Billing), y=Average_Billing)) +
  geom_bar(stat="identity", fill="blue", alpha=0.7) +
  theme_minimal() +
  ggtitle("Average Billing Amount by Insurance Provider") +
  coord_flip()
```

## Almost all patient across different insurance providers were charged the same amount in terms of medical costs.