---
title: "HYPERTENSION PREVALENCE STUDY"
author: "Enock Bereka"
date: "2025-01-30"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load necessary libraries
```{r warning = FALSE, message=FALSE}
library(tidyverse)
library(performance)
library(ggeffects)
library(sjPlot)
library(gtsummary)
library(vip)
library(pROC)
library(car)
library(ggcorrplot)
```

## Load and inspect the dataset
```{r warning = FALSE, message=FALSE}
data <- read_csv("C:/Users/PC/OneDrive/Desktop/Data Science/Datasets/Datasets/Hypertension_data.csv")
glimpse(data)
anyNA(data)
duplicated(data) %>% table()
```

## Data Preparation
## Assign ID to act as a unique identifier
```{r warning = FALSE, message=FALSE}
data$id <- seq.int(nrow(data))
data <- unique(data)
```

## Convert Categorical variables to factors
```{r warning = FALSE, message=FALSE}
data$education <- as.factor(data$education)
```

## Categorise Age and BMI into groups
```{r warning = FALSE, message=FALSE}
data$bmi <- ifelse(data$BMI<18.5, "Underweight",
                       ifelse(data$BMI>=18.5&data$BMI<25, "Normal Weight",                         ifelse(data$BMI>=25&data$BMI<30, "Overweight",                           ifelse(data$BMI>=30&data$BMI<40, "Obese", "Severely Obese"))))

data$bmi <- factor(data$bmi, levels = c("Underweight", "Normal Weight", "Overweight", 
"Obese", "Severely Obese"))

data$Age <- ifelse(data$age <= 50, "Middle_age", "Old_age")

data$Hypertension_prevalence <- ifelse(data$prevalentHyp == 0, "Negative", "Positive")
data$Smoking_status <- ifelse(data$currentSmoker == 1, "Non Smoker", "Smoker")
```

## Univariate analysis
## Summary statistics for continous columns
```{r warning = FALSE, message=FALSE}
be <- c("age","BMI","heartRate")
summary(data[be])
```

## Frequency distributions
## Continous columns
## Theme set
```{r warning = FALSE, message=FALSE}
theme_set(theme_classic()+theme(title = element_text(face = "bold", colour = "steelblue")))
```

## Age
```{r warning = FALSE, message=FALSE}
ggplot(data, aes(x = age))+
  geom_histogram(bins = 15, fill = "violet", color = "green")+
  labs(title = "Age Distributions", y = "Frequency")

ggplot(data, aes(x = age))+
  geom_density(fill = "violet", color = "green")+
  labs(title = "Age Distributions", y = "Frequency")
```

## BMI
```{r warning = FALSE, message=FALSE}
ggplot(data, aes(x = BMI))+
  geom_histogram(bins = 15, fill = "violet", color = "green")+
  labs(title = "BMI Distributions", y = "Frequency")

ggplot(data, aes(x = BMI))+
  geom_density(fill = "violet", color = "green")+
  labs(title = "BMI Distributions", y = "Frequency")
```

## Heart Rate
```{r warning = FALSE, message=FALSE}
ggplot(data, aes(x = heartRate))+
  geom_histogram(bins = 15, fill = "violet", color = "green")+
  labs(title = "Heart Rate Distributions", y = "Frequency")

ggplot(data, aes(x = heartRate))+
  geom_density(fill = "violet", color = "green")+
  labs(title = "Heart Rate Distributions", y = "Frequency")
```

## Categorical variables
## Education
```{r warning = FALSE, message=FALSE}
ggplot(data, aes(x = education, fill = education))+
  geom_bar()+
  labs(title = "Distribution of participants across education",
       y = "Frequency")
```

## Hypertension Prevalence
```{r warning = FALSE, message=FALSE}
ggplot(data, aes(x = Hypertension_prevalence, fill = Hypertension_prevalence))+
  geom_bar()+
  labs(title = "Hypertension Prevalence",
       y = "Frequency")
```

## Smoking Status
```{r warning = FALSE, message=FALSE}
ggplot(data, aes(x = Smoking_status, fill = Smoking_status))+
  geom_bar()+
  labs(title = "Smoking Status Distribution",
       y = "Frequency")
```

## Age group
```{r warning = FALSE, message=FALSE}
ggplot(data, aes(x = Age, fill = Age))+
  geom_bar()+
  labs(title = "Age group Distribution",
       y = "Frequency")
```
## Majority of the participants were middle aged
## BMI Levels
```{r warning = FALSE, message=FALSE}
ggplot(data, aes(x = bmi, fill = bmi))+
  geom_bar()+
  labs(title = "BMI levels Distribution",
       y = "Frequency")
```

## Majority of the participants had normal and overweight

## Bivariate analysis
## Hypothesis testing
## Correlations
```{r warning = FALSE, message=FALSE}
cor(data[be])
```

## Visualize the correlations
```{r warning = FALSE, message=FALSE}
cr <- cor(data[be])
ggcorrplot(cr,title = "Correlation Heatmap",lab_col = "black",
           lab = TRUE, legend.title = "Pearson Correlation",
           lab_size = 3, ggtheme = theme_classic(),
           outline.color = "black",
           colors = c("orange", "green", "blue"))
```

## There is a positive weak correlation between age and BMI
## There is no correlation between age and heart rate
## There is a weak positive correlation between BMI and heart rate

## Chi-square
## What is the relationship between smoking status and hypertension prevalence
```{r warning = FALSE, message=FALSE}
contingency_table <- table(data$Hypertension_prevalence, data$Smoking_status)
contingency_table 
chisq_result <- chisq.test(contingency_table)
print(chisq_result)
```

## There is an a statistically significant association between hypertension prevalence and smoking status indicated by a lower significant p-value and higher chi-square statistic value.

## Visualize your chisq result
```{r warning = FALSE, message=FALSE}
ggplot(data, aes(Smoking_status, fill = Hypertension_prevalence))+
  geom_bar(position = "dodge")+
  labs(title = "Hypertension Prevalence based on smoking status",
       y = "Frequency")
```

## What is the relationship between education levels and hypertension prevalence
```{r warning = FALSE, message=FALSE}
contingency_table <- table(data$Hypertension_prevalence, data$education)
contingency_table 
chisq_result <- chisq.test(contingency_table)
print(chisq_result)
```

## There is no significant association between hypertension prevalence and education levels indicated by p-value above 0.05

## Visualize your chisq result
```{r warning = FALSE, message=FALSE}
ggplot(data, aes(education, fill = Hypertension_prevalence))+
  geom_bar(position = "dodge")+
  labs(title = "Hypertension Prevalence based on Education status",
       y = "Frequency")
```

## What is the relation between age groups and hypertension prevalence
```{r warning = FALSE, message=FALSE}
contingency_table <- table(data$Hypertension_prevalence, data$Age)
contingency_table 
chisq_result <- chisq.test(contingency_table)
print(chisq_result)
```

## There is a statistically significant association between hypertension prevalence and age groups as shown by a p-value less than 0.05.

## Visualize your chisq result
```{r warning = FALSE, message=FALSE}
ggplot(data, aes(Age, fill = Hypertension_prevalence))+
  geom_bar(position = "dodge")+
  labs(title = "Hypertension Prevalence based on Age groups",
       y = "Frequency")
```

## What is the relationship between BMI levels and hypertension prevalence
```{r warning = FALSE, message=FALSE}
contingency_table <- table(data$Hypertension_prevalence, data$bmi)
contingency_table 
chisq_result <- chisq.test(contingency_table)
print(chisq_result)
```

## There is a statistically significant association between hypertension prevalence and BMI levels indicated by a p-value less than 0.05.

## Visualize your chisq result
```{r warning = FALSE, message=FALSE}
ggplot(data, aes(bmi, fill = Hypertension_prevalence))+
  geom_bar(position = "dodge")+
  labs(title = "Hypertension Prevalence based on BMI levels",
       y = "Frequency")
```

## T-test
## What is the relationship between Age and Hypertension Prevalence
```{r warning = FALSE, message=FALSE}
t.test_result <- t.test(data$age, data$prevalentHyp)
print(t.test_result)
```

## True difference in means of age across hypertension status is not equal to zero indicating a significant association.

## Visualize the t.test_result
```{r warning = FALSE, message=FALSE}
ggplot(data, aes(x = Hypertension_prevalence, y = age , fill = Hypertension_prevalence))+
  geom_boxplot()+
  labs(title = "Hypertension Prevalence based on Age")
```

## What is the relationship between BMI and Hypertension Prevalence
```{r warning = FALSE, message=FALSE}
t.test_result <- t.test(data$BMI, data$prevalentHyp)
print(t.test_result)
```

## True difference in means of BMI across hypertension status is not equal to zero indicating a significant association.

## Visualize the t.test_result
```{r warning = FALSE, message=FALSE}
ggplot(data, aes(x = Hypertension_prevalence, y = BMI , fill = Hypertension_prevalence))+
  geom_boxplot()+
  labs(title = "Hypertension Prevalence based on BMI")
```

## What is the relationship between Heart Rate and Hypertension Prevalence
```{r warning = FALSE, message=FALSE}
t.test_result <- t.test(data$heartRate, data$prevalentHyp)
print(t.test_result)
```

## True difference in means of heart rate across hypertension status is not equal to zero indicating a significant association.

## Visualize the t.test_result
```{r warning = FALSE, message=FALSE}
ggplot(data, aes(x = Hypertension_prevalence, y = heartRate , fill = Hypertension_prevalence))+
  geom_boxplot()+
  labs(title = "Hypertension Prevalence based on Heart Rate")
```

## ANOVA
## How does Age vary across BMI levels
```{r warning = FALSE, message=FALSE}
anova_result <- aov(age~bmi, data = data)
summary(anova_result)
```

## The p-value associated with BMI levels is very small indicating a strong statistical significance. This suggests that there is a statistically significant difference in age across different BMI levels.

## Visualize the anova_result
```{r warning = FALSE, message=FALSE}
ggplot(data, aes(x = bmi, y = age, fill = bmi))+
  geom_boxplot()+
  labs(title = "Age differences across BMI levels")
```

## How does Heart rate vary across BMI levels
```{r warning = FALSE, message=FALSE}
anova_result <- aov(heartRate~bmi, data = data)
summary(anova_result)
```

## The p-value associated with BMI levels is very small indicating a strong statistical significance. This suggests that there is a statistically significant difference in heart rate across different BMI levels.

## Visualize the anova_result
```{r warning = FALSE, message=FALSE}
ggplot(data, aes(x = bmi, y = heartRate, fill = bmi))+
  geom_boxplot()+
  labs(title = "Heart rate differences across BMI levels")
```

## How does BMI vary across education status
```{r warning = FALSE, message=FALSE}
anova_result <- aov(BMI~education, data = data)
summary(anova_result)
```

## The p-value(0.774) is substantially higher than the common significance value of 0.05. This indicates there isnt a statistically significant difference in BMI across education levels. 

## Visualize the anova_result
```{r warning = FALSE, message=FALSE}
ggplot(data, aes(x = education, y = BMI, fill = education))+
  geom_boxplot()+
  labs(title = "BMI differences across education levels")
```

## How does Age vary across education status
```{r warning = FALSE, message=FALSE}
anova_result <- aov(age~education, data = data)
summary(anova_result)
```

## The p-value(0.781) is substantially higher than the common significance value of 0.05. This indicates there isnt a statistically significant difference in Age across education levels.

## Visualize the anova_result
```{r warning = FALSE, message=FALSE}
ggplot(data, aes(x = education, y = age, fill = education))+
  geom_boxplot()+
  labs(title = "Age differences across education levels")
```

## How does Heart Rate vary across education status
```{r warning = FALSE, message=FALSE}
anova_result <- aov(heartRate~education, data = data)
summary(anova_result)
```

## The p-value(0.638) is substantially higher than the common significance value of 0.05. This indicates there isnt a statistically significant difference in BMI across education levels.

## Visualize the anova_result
```{r warning = FALSE, message=FALSE}
ggplot(data, aes(x = education, y = heartRate , fill = education))+
  geom_boxplot()+
  labs(title = "Heart Rate differences across education levels")
```

## Multivariate analysis
## Logistic Regression
```{r warning = FALSE, message=FALSE}
original <- subset(data, select = c("education", "age", "BMI", "currentSmoker","prevalentHyp", "heartRate"))
```

## Set up factors
```{r warning = FALSE, message=FALSE}
original$currentSmoker <- as.factor(original$currentSmoker)
original$prevalentHyp <- as.factor(original$prevalentHyp)
```

## Implement the logistic regression
```{r warning = FALSE, message=FALSE}
model <- glm(prevalentHyp~., data = original, family = "binomial")
```

## Check model assumptions
```{r warning = FALSE, message=FALSE}
check_predictions(model) %>% plot()
check_outliers(model)
check_outliers(model) %>% plot()
check_collinearity(model)
check_residuals(model)
check_residuals(model) %>% plot()
```


## Check model predicted probabilities
```{r warning = FALSE, message=FALSE}
ggeffect(model)
```

## Visualize model predictions
```{r warning = FALSE, message=FALSE}
ggpredict(model) %>% plot() %>% 
  sjPlot::plot_grid()
```

## Hypertension prevalence increases with increase in Age
## Hypertension prevalence increases with increase in bmi
## Hypertension prevalence increases with increase in heart rate
## Smokers are at the high risk of developing hypertension compared to non-smokers.
## Hypertension prevalence decreases as the education levels increases.

## Check for odds ratios
```{r warning = FALSE, message=FALSE}
tbl_regression(
  model, exponentiate = TRUE,
  add_pairwise_contrasts = TRUE,
  contrasts_adjust = "bonferroni",
  pairwise_reverse = FALSE,
  pvalue_fun = ~style_pvalue(.x, digits = 3)) %>% add_significance_stars(hide_p = F, hide_se = T,
                         hide_ci = F) %>% 
  bold_p()
```

## For every one unit change in age, there is 1.08 change in odds of developing hypertension.
## For every one unit change in BMI, there is 1.16 change in odds of developing hypertension.
## For every one unit change in heart rate, there is 1.03 change in odds of developing hypertension.

## Checking variable importance
```{r warning = FALSE, message=FALSE}
car::Anova(model)
vip(model)
```

## Based on this study, Age, BMI and Heart rate are the top predictor of hypertension prevalence.

## How accurate is our model in making predictions
## Produce ROC curve
```{r warning = FALSE, message=FALSE}
roc(prevalentHyp~fitted.values(model), data = original,plot = TRUE, legacy.axes = TRUE, 
print.auc = TRUE, ci = TRUE)
```

## The model achieved an AUC of 0.762 making it a valuable tool for making predictions.