---
title: "MALARIA ANALYSIS REPORT"
author: "Enock Bereka"
date: "2025-01-26"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load necessary libraries
```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(performance)
library(factoextra)
library(vip)
library(sjPlot)
library(pROC)
library(ggeffects)
library(gtsummary)
library(car)
library(ggcorrplot)
library(cluster)
```

## Load and inspect the dataset
```{r warning=FALSE, message=FALSE}
data <- read_csv("C:/Users/PC/OneDrive/Desktop/Data Science/Datasets/Datasets/Mal neg & asymptomatic.csv")
glimpse(data)
anyNA(data)
duplicated(data) %>% table()
```

## Clean the data: Remove empty columns
```{r warning=FALSE, message=FALSE}
data$...4 <- NULL
data$...6 <- NULL
data$...8 <- NULL
```

## Convert Participants to factors
```{r warning=FALSE, message=FALSE}
data$Participants <- as.factor(data$Participants)
date = data
```

## Univariate Analysis
## Descriptive Statistics for continous columns
```{r warning=FALSE, message=FALSE}
dat <- data %>% select(-1)
dat %>% summary()
```

## Frequency Distributions
```{r warning=FALSE, message=FALSE}
theme_set(theme_classic()+
            theme(title = element_text(face = "bold",
                                       colour = "darkred")))
```

## Participants
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(Participants))+
  geom_bar(fill = "green", color = "black")+
  labs(title = "Distribution of participants")
```

## Mature neutrophils
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(`Mature (CD11b+ CD62L+ CD182+ CD184-)`))+
  geom_histogram(aes(y = ..density..),bins = 6, fill = "green", color = "black")+
  geom_density(color = "red", linewidth = 1)+
  labs(title = "Distribution of mature neutrophils", x = "Mature neutophils")
```

## Unconventional neutrophils
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(`Unconventional Aged(CD62L+ CD184+)`))+
  geom_histogram(aes(y = ..density..), bins = 6, fill = "green", color = "black")+
  geom_density(color = "red", linewidth = 1)+
  labs(title = "Distribution of unconventional neutrophils", x = "Unconventional neutophils")
```

## Aged neutrophils
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(`Aged(CD62L- CD184+)`))+
  geom_histogram(aes(y = ..density..), bins = 4, fill = "green", color = "black")+
  geom_density(color = "red", linewidth = 1)+
  labs(title = "Distribution of aged neutrophils", x = "Aged neutophils")
```

## Mature inactive neutrophils
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(`Mature Inactive(CD11b- CD62L+ CD182+ CD184-)`))+
  geom_histogram(aes(y = ..density..), bins = 2,width = 0.5, fill = "green", color = "black")+
  geom_density(color = "red", linewidth = 1)+
  labs(title = "Distribution of Mature inactive neutrophils", x = "Mature inactive neutrophils")
```

## Immature neutrophils
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(`Immature(CD10-)`))+
  geom_histogram(aes(y = ..density..), bins = 5,width = 0.5, fill = "green", color = "black")+
  geom_density(color = "red", linewidth = 1)+
  labs(title = "Distribution of Immature neutrophils", x = "Immature neutrophils")
```

## Bivariate Analysis
## Correlation
## Relationship between Continous variables
## Correlation Analysis
```{r warning=FALSE, message=FALSE}
corr_matrix <- dat %>%
  cor(use = "pairwise.complete.obs")
```

## Correlation heatmap
```{r warning=FALSE, message=FALSE}
ggcorrplot(corr_matrix, lab = TRUE, title = "Correlation Matrix of Immune Cells")
```

# Differences in immune profiles across participants groups
```{r warning=FALSE, message=FALSE}
data_long <- data %>%
  pivot_longer(cols = -Participants, names_to = "Immune_Population", values_to = "Value")
ggplot(data_long, aes(x = Participants, y = Value, fill = Participants)) +
  geom_boxplot() +
  facet_wrap(~ Immune_Population, scales = "free") +
  theme_minimal() +
  labs(title = "Immune Profiles by Group", x = "Group", y = "Value")
```

## Multivariate Analysis
## Clustering Analysis
## Scale numeric columns for clustering
```{r warning=FALSE, message=FALSE}
scaled_data <- data %>%
  select(where(is.numeric)) %>%
  scale()
```

## Perform k-means clustering
```{r warning=FALSE, message=FALSE}
set.seed(123) # for reproducibility
kmeans_result <- kmeans(scaled_data, centers = 3) # Adjust the number of centers as needed
```
```{r warning=FALSE, message=FALSE}
## Add cluster assignments to the original data
data$Cluster <- as.factor(kmeans_result$cluster)
```

## Visualize clustering results using PCA
```{r warning=FALSE, message=FALSE}
fviz_cluster(kmeans_result, data = scaled_data, geom = "point", stand = FALSE)
```

## Logistic Regression
## Rename the columns
```{r warning=FALSE, message=FALSE}
date <- rename(date,
               "Mature" = `Mature (CD11b+ CD62L+ CD182+ CD184-)`,
               "Unconventional" = `Unconventional Aged(CD62L+ CD184+)`,
               "Aged" = `Aged(CD62L- CD184+)`,
               "Mature_inactive" = `Mature Inactive(CD11b- CD62L+ CD182+ CD184-)`,
               "Immature" = `Immature(CD10-)`)
date$Participants <- ifelse(date$Participants=="asymptomatic", 0, 1)
date$Participants <- as.factor(date$Participants)
```

## Implement the model
```{r warning=FALSE, message=FALSE}
model <- glm(Participants~., data = date, family = binomial)
```

## Check Model assumptions
```{r warning=FALSE, message=FALSE}
check_predictions(model) %>% plot()
check_collinearity(model)
check_residuals(model)
```

## Check model Predictions
```{r warning=FALSE, message=FALSE}
ggeffect(model)
```

## Visualize model predictions
```{r warning=FALSE, message=FALSE}
ggeffect(model) %>% plot() %>% 
  sjPlot::plot_grid()
```

## Display odds ratios
```{r warning=FALSE, message=FALSE}
tbl_regression(
  model, exponentiate = TRUE,
  add_pairwise_contrasts = TRUE,
  contrasts_adjust = "bonferroni",
  pairwise_reverse = FALSE,
  pvalue_fun = ~style_pvalue(.x, digits = 3)) %>% 
  add_significance_stars(hide_p = F, hide_se = T,
                         hide_ci = F) %>% 
  bold_p()
```

## Check variable importance
```{r warning=FALSE, message=FALSE}
car::Anova(model)
vip(model)
```

#Produce ROC curve
```{r warning=FALSE, message=FALSE}
roc(Participants~fitted.values(model), data = date,
    plot = TRUE, legacy.axes = TRUE, 
    print.auc = TRUE, ci = TRUE)
```

