---
title: "MEDICAL DATA ANALYTICS CALORIES DETERMINATION"
author: "Enock Bereka"
date: "2025-02-15"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load the necessary libraries
```{r warning=FALSE, message=FALSE}
library(vip)
library(performance)
library(ggeffects)
library(gtsummary)
library(equatiomatic)
library(tidyverse)
library(ggcorrplot)
library(sjPlot)
library(emmeans)
library(effectsize)
library(car)
library(vip)
```

## Load and inspect the dataset
```{r warning=FALSE, message=FALSE}
data <- read_csv("C:/Users/PC/OneDrive/Desktop/Data Science/Datasets/Datasets/calories.csv")
glimpse(data)
anyNA(data)
sum(duplicated(data))
```

## Data preparation
```{r warning=FALSE, message=FALSE}
data$Gender <- as.factor(data$Gender)
```

## Univariate analysis
## Summary statistics
```{r warning=FALSE, message=FALSE}
data %>% select(-1, -2) %>% summary()
```

## Frequency distributions
```{r warning=FALSE, message=FALSE}
theme_set(theme_classic()+theme(title = element_text
                                (face = "bold", 
                                  colour = "green")))
```

## Gender
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(Gender))+
  geom_bar(fill = "purple", color = "red")+
  labs(title = "Distribution of subjects based on Gender",
       y = "Frequency", caption = "enoch(2025)")
```

## Age
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(Age))+
  geom_histogram(fill = "purple", color = "red", bins = 15)+
  labs(title = "Distribution of subjects based on Age",
       y = "Frequency", caption = "enoch(2025)")
```

## Height
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(Height))+
  geom_histogram(fill = "purple", color = "red", bins = 15)+
  labs(title = "Distribution of subjects based on Height",
       y = "Frequency", caption = "enoch(2025)")
```

## Weight
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(Weight))+
  geom_histogram(fill = "purple", color = "red", bins = 15)+
  labs(title = "Distribution of subjects based on Weight",
       y = "Frequency", caption = "enoch(2025)")
```

## Duration
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(Duration))+
  geom_histogram(fill = "purple", color = "red", bins = 15)+
  labs(title = "Distribution of subjects based on Duration",
       y = "Frequency", caption = "enoch(2025)")
```

## Heart Rate
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(Heart_Rate))+
  geom_histogram(fill = "purple", color = "red", bins = 15)+
  labs(title = "Distribution of subjects based on Heart Rate",
       y = "Frequency", caption = "enoch(2025)",
       x = "Heart Rate")
```

## Body Temperature
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(Body_Temp))+
  geom_histogram(fill = "purple", color = "red", bins = 15)+
  labs(title = "Distribution of subjects based on Body Temperature",
       y = "Frequency", caption = "enoch(2025)",
       x = "Body Temperature")
```

## Calories
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(Calories))+
  geom_histogram(fill = "purple", color = "red", bins = 15)+
  labs(title = "Distribution of subjects based on Calories",
       y = "Frequency", caption = "enoch(2025)")
```

## Bivariate Analysis
## Correlation Analysis
```{r warning=FALSE, message=FALSE}
cr <- data %>% select(-1, -2) %>% cor()
ggcorrplot(cr,title = "Correlation Heatmap",lab_col = "black",
           lab = TRUE, legend.title = "Pearson Correlation",
           lab_size = 3, ggtheme = theme_classic(),
           outline.color = "black",
           colors = c("orange", "green", "blue"))
```

## There is a strong positive correlation between Body temperature, heart rate and duration with calories.
## There is a strong correlation between height and weight indicating multicollinearity.

## How Calories relate with Gender
```{r warning=FALSE, message=FALSE}
t.test(Calories~Gender, data = data)
``` 

## There is a statistically significant difference in mean of calories between male and female indicated by smaller p-value lower than 0.05. This shows that gender has some influence on calories determination.

## Visualize the relationship
```{r warning=FALSE, message=FALSE}
ggplot(data, aes(x = Gender, y = Calories))+
  geom_boxplot(fill = "purple", color = "red")+
  labs(title = "Calories vs Gender", caption = "enoch(2025)")
```

## Multivariate analysis
## Linear Regression
```{r warning=FALSE, message=FALSE}
data$User_ID <- NULL
model <- lm(Calories~., data = data)
```

## Check Model assumptions
## Check for outliers
```{r warning=FALSE, message=FALSE}
check_outliers(model)
check_outliers(model) %>% plot()
```

## Check for autocorrelation
```{r warning=FALSE, message=FALSE}
check_autocorrelation(model)
```

## Check residuals
```{r warning=FALSE, message=FALSE}
plot_model(model, type = "diag")
```

## Visualizing Model Diagnostics
## Residual Plot
```{r warning=FALSE, message=FALSE}
plot(model, which = 1, col = "violet")  # Residuals vs. Fitted plot
```

## Normal Q-Q Plot
```{r warning=FALSE, message=FALSE}
plot(model, which = 2, col = "red", main = "Normal Q-Q Plot")  # Normal Q-Q plot
```

## Cook’s Distance Plot (Influence)
```{r warning=FALSE, message=FALSE}
plot(model, which = 4, col = "violet")
```

## Determine model predicted probabilities
```{r warning=FALSE, message=FALSE}
ggeffect(model)
```

## Marginal Effects of Gender on Calories
```{r warning=FALSE, message=FALSE}
plot_model(model, type = "pred", terms = "Gender")
```

## Females have higher volumes of calories than females

## Marginal Effects of Age on Calories
```{r warning=FALSE, message=FALSE}
plot_model(model, type = "pred", terms = "Age") 
```

## An increase in Age leads to an increase in calories.


## Marginal Effects of Height on Calories
```{r warning=FALSE, message=FALSE}
plot_model(model, type = "pred", terms = "Height")
```

## An increase in height leads to a decrease in calories.

## Marginal Effects of Weight on Calories
```{r warning=FALSE, message=FALSE}
plot_model(model, type = "pred", terms = "Weight")
```

## An increase in weight leads to an increase in calories.

## Marginal Effects of Duration on Calories
```{r warning=FALSE, message=FALSE}
plot_model(model, type = "pred", terms = "Duration")
```

## An increase in duration leads to an increase in calories.

## Marginal Effects of Heart Rate on Calories
```{r warning=FALSE, message=FALSE}
plot_model(model, type = "pred", terms = "Heart_Rate")
```

## An increase in heart rate leads to an increase in calories.

## Marginal Effects of Body Temperature on Calories
```{r warning=FALSE, message=FALSE}
plot_model(model, type = "pred", terms = "Body_Temp")
```

## An increase in body temperature leads to decrease i calories.

## Model interactions
## Fit a model with an interaction term(Gender)
```{r warning=FALSE, message=FALSE}
model_interaction <- lm(Calories ~ Age * Gender, data = data)
```

## Visualize interaction effect
```{r warning=FALSE, message=FALSE}
plot_model(model_interaction, type = "int")
```

## Fit a model with an interaction term(Gender)
```{r warning=FALSE, message=FALSE}
model_interaction <- lm(Calories ~ Height * Gender, data = data)
```

## Visualize interaction effect
```{r warning=FALSE, message=FALSE}
plot_model(model_interaction, type = "int")
```

## Fit a model with an interaction term(Gender)
```{r warning=FALSE, message=FALSE}
model_interaction <- lm(Calories ~ Weight * Gender, data = data)
```

## Visualize interaction effect
```{r warning=FALSE, message=FALSE}
plot_model(model_interaction, type = "int")
```

## Fit a model with an interaction term(Gender)
```{r warning=FALSE, message=FALSE}
model_interaction <- lm(Calories ~ Duration * Gender, data = data)
```

## Visualize interaction effect
```{r warning=FALSE, message=FALSE}
plot_model(model_interaction, type = "int")
```

## Fit a model with an interaction term(Gender)
```{r warning=FALSE, message=FALSE}
model_interaction <- lm(Calories ~ Body_Temp * Gender, data = data)
```

## Visualize interaction effect
```{r warning=FALSE, message=FALSE}
plot_model(model_interaction, type = "int")
```

## Fit a model with an interaction term(Gender)
```{r warning=FALSE, message=FALSE}
model_interaction <- lm(Calories ~ Heart_Rate * Gender, data = data)
```

## Visualize interaction effect
```{r warning=FALSE, message=FALSE}
plot_model(model_interaction, type = "int")
```

## Obtaining Estimated Marginal Means
```{r warning=FALSE, message=FALSE}
emmeans(model, ~ Gender)
```

## Pairwise Comparisons of Estimated Means
```{r warning=FALSE, message=FALSE}
pairs(emmeans(model, ~ Gender))
```

## Visualizing Estimated Marginal Means
## Basic Visualization
```{r warning=FALSE, message=FALSE}
emmip(model, Gender ~ Age)
```

## Compare relative importance of predictors
```{r warning=FALSE, message=FALSE}
plot_model(model, type = "std")
```

## Generate a regression summary table
```{r warning=FALSE, message=FALSE}
tbl_regression(model)
```

## Produce model equation
```{r warning=FALSE, message=FALSE}
extract_eq(model)
```

## Checking variable importance
```{r warning=FALSE, message=FALSE}
car::Anova(model)
vip(model)
```

## How well our model fits the data
```{r warning=FALSE, message=FALSE}
performance(model)
```

## Our model explains 96.7% of variability in the outcome variable making it a better fit.
## Our model has RMSE of 11.311 which is less than the standard deviation of the outcome variable making it an excellent tool in making predictions.