---
title: "POISSON AND NEGATIVE BINOMIAL REGRESSION MODELS"
author: "Enock Bereka"
date: "2025-02-14"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Load the required packages
```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(MASS)
```

## Load and inspect the data
```{r warning=FALSE, message=FALSE}
data("epil")
glimpse(epil)
anyNA(epil)
```

## Summary of the data
```{r warning=FALSE, message=FALSE}
summary(epil)
```

## Visualize the outcome variable
## Histogram of seizure counts
```{r warning=FALSE, message=FALSE}
ggplot(epil, aes(x = y)) + 
  geom_histogram(bins = 15, fill = "blue", color = "black") +
  labs(title = "Distribution of Seizure Counts", x = "Number of Seizures", y = "Frequency") +
  theme_minimal()
```

## Fit a Poisson regression model
```{r warning=FALSE, message=FALSE}
poisson_model <- glm(y ~ trt + age + lbase + offset(log(period)), 
                     family = poisson, data = epil)
```

## View model results
```{r warning=FALSE, message=FALSE}
summary(poisson_model)
```

## Check for overdispersion
## Compute dispersion ratio
```{r warning=FALSE, message=FALSE}
dispersion_ratio <- sum(residuals(poisson_model, type = "pearson")^2) / poisson_model$df.residual
dispersion_ratio
```

## The poisson model has a dispersion ratio of 9.7 indicating overdispersion.

## Addressing Overdispersion with Negative Binomial Regression
## Fit Negative Binomial Model
```{r warning=FALSE, message=FALSE}
nb_model <- glm.nb(y ~ trt + age + lbase + offset(log(period)), data = epil)
```

## Summary of Negative Binomial Model
```{r warning=FALSE, message=FALSE}
summary(nb_model)
```
## Treatment and lod counts for baseline period are significant in the negaitve binomial model indicating they can be used to predict seizure counts for Epileptics.

## Model comparison
## Compare AIC values
```{r warning=FALSE, message=FALSE}
AIC(poisson_model, nb_model)
```

## We choose negative binomial model since it has lower AIC compared to poisson model.

## Plot actual vs predicted values
```{r warning=FALSE, message=FALSE}
epil$predicted <- predict(nb_model, type = "response")

ggplot(epil, aes(x = predicted, y = y)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "red") +
  labs(title = "Predicted vs. Actual Seizure Counts", x = "Predicted Counts", y = "Actual Counts") +
  theme_minimal()
```

## Create new data for prediction
```{r warning=FALSE, message=FALSE}
new_data <- data.frame(trt = c(1, 0), age = c(30, 40), lbase = c(2, 3), period = c(10, 10))
new_data$trt <- factor(new_data$trt, levels = c(1, 0), labels = c("progabide", "placebo"))
```

## Predict seizure counts
```{r warning=FALSE, message=FALSE}
new_data$predicted_seizures <- predict(nb_model, newdata = new_data, type = "response")
```

## View predictions
```{r warning=FALSE, message=FALSE}
new_data
```

## Checking assumptions
## Residual Plot
```{r warning=FALSE, message=FALSE}
ggplot(epil, aes(x = predicted, y = residuals(poisson_model, type = "pearson"))) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Residuals vs. Predicted Values", x = "Predicted Counts", y = "Pearson Residuals") +
  theme_minimal()
```

