---
title: "TB MORTALITY TRENDS IN SUB SAHARAN AFRICA"
author: "Enock Bereka"
date: "2025-02-14"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Required Libraries
```{r warning=FALSE, message=FALSE}
library(tidyverse)    # Data manipulation and visualization
library(Epi)          # Age-Period-Cohort modeling
library(apc)          # Alternative APC analysis
library(ggplot2)      # Additional plotting functions
library(viridis)      # Color palette for heatmaps
library(MASS)         # For glm.nb (negative binomial regression if overdispersion is present)
```

## Load and inspect the Data
```{r warning=FALSE, message=FALSE}
tb_data <-  read_csv("C:/Users/PC/OneDrive/Desktop/Data Science/Datasets/Datasets/simulated_tb_mortality_SSA_no_cohort.csv")
glimpse(tb_data)
anyNA(tb_data)
sum(duplicated(tb_data))
```

## Compute the Cohort Variable
```{r warning=FALSE, message=FALSE}
tb_data <- tb_data %>%
  mutate(Age = factor(Age, ordered = TRUE), # Convert Age to ordered factor
         Period = as.numeric(Period),      # Ensure Period is numeric
         Cohort = Period - as.numeric(Age)) # Compute Cohort
```

## Exploratory Data Analysis
```{r warning=FALSE, message=FALSE}
summary(tb_data)
ggplot(tb_data, aes(x = as.numeric(Age), y = Mortality_Rate))+
  geom_point(alpha = 0.5, color = "red")+
  geom_smooth(method = "loess")+
  theme_classic()+
  labs(title = "Age vs Mortality Rates in Sub Saharan Africa", x = "Age",
       y = "Mortality Rate",
       caption = "TB Mortality Sub Saharan Africa(1990-2020)")
```

## Visualizing Mortality Trends Over Time
```{r warning=FALSE, message=FALSE}
ggplot(tb_data, aes(x = Period, y = Mortality_Rate))+
  geom_line(linewidth = 1.1, color = "red")+
  theme_minimal()+
  geom_smooth(method = "lm")+
  labs(title = "TB Mortality Trends in Sub Saharan Africa ",
       x = "Year",
       y = "Mortality Rate per 100,000",
       caption = "TB Mortality Sub Saharan Africa(1990-2020)")
```

## Visualizing Mortality Trends Over Time across Age groups
```{r warning=FALSE, message=FALSE}
ggplot(tb_data, aes(x = Period, y = Mortality_Rate, color = as.factor(Age))) +
  geom_line(linewidth = 1.1) +
  theme_minimal() +
  labs(title = "TB Mortality Trends in Sub Saharan Africa across Age groups",
       x = "Year",
       y = "Mortality Rate per 100,000",
       color = "Age Group",
       caption = "TB Mortality Sub Saharan Africa(1990-2020)")
```

## Visualizing Mortality Trends across Cohorts
```{r warning=FALSE, message=FALSE}
ggplot(tb_data, aes(x = Cohort, y = Mortality_Rate))+
  geom_line(linewidth = 1, color = "red")+
  theme_minimal()+
  geom_smooth(method = "lm")+
  labs(title = "TB Mortality Trends in Sub Saharan Africa across Cohorts",
       x = "Birth Year",
       y = "Mortality Rate per 100,000",
       caption = "TB Mortality Sub Saharan Africa(1990-2020)")
```

## Age-Period-Cohort Model using Poisson Regression
```{r warning=FALSE, message=FALSE}
apc_model <- glm(Mortality_Rate ~ factor(Age) + factor(Period) + factor(Cohort), 
                 data = tb_data, 
                 family = poisson)
```

## Checking for Overdispersion 
```{r warning=FALSE, message=FALSE}
dispersion <- sum(residuals(apc_model, type = "pearson")^2) / df.residual(apc_model)
```

## If dispersion is detectes i use Negative Binomial Model
```{r warning=FALSE, message=FALSE}
if (dispersion > 1.5) {
  cat("Overdispersion detected (dispersion =", dispersion, "). Using Negative Binomial Regression.\n")
  apc_model <- glm.nb(Mortality_Rate ~ factor(Age) + factor(Period) + factor(Cohort), 
                      data = tb_data)
  summary(apc_model)
}
```

## Visualizing Age, Period, and Cohort Effects
## Extract and plot the APC effects

## Age Effects
```{r warning=FALSE, message=FALSE}
age_effects <- coef(apc_model)[grep("Age", names(coef(apc_model)))]
age_df <- data.frame(Age = levels(tb_data$Age)[-1], Effect = age_effects)
ggplot(age_df, aes(x = Age, y = Effect)) +
  geom_point(size = 3, color = "blue") +
  geom_line(group = 1, color = "blue", linetype = "dashed") +
  theme_minimal() +
  labs(title = "Age Effect on TB Mortality in Sub Saharan Africa", x = "Age Group", y = "Effect Size",
       caption = "TB Mortality Sub Saharan Africa(1990-2020)")
```

## Period Effects
```{r warning=FALSE, message=FALSE}
period_effects <- coef(apc_model)[grep("Period", names(coef(apc_model)))]
period_df <- data.frame(Period = unique(tb_data$Period)[-1], Effect = period_effects)
ggplot(period_df, aes(x = Period, y = Effect)) +
  geom_point(size = 3, color = "red") +
  geom_line(group = 1, color = "red", linetype = "dashed") +
  theme_minimal() +
  labs(title = "Period Effect on TB Mortality in Sub Saharan Africa", x = "Period", y = "Effect Size",
       caption = "TB Mortality Sub Saharan Africa(1990-2020)")
```

## Cohort Effects
```{r warning=FALSE, message=FALSE}
cohort_effects <- coef(apc_model)[grep("Cohort", names(coef(apc_model)))]
cohort_df <- data.frame(Cohort = unique(tb_data$Cohort)[-1], Effect = cohort_effects)
ggplot(cohort_df, aes(x = Cohort, y = Effect)) +
  geom_point(size = 3, color = "green") +
  geom_line(group = 1, color = "green", linetype = "dashed") +
  theme_minimal() +
  labs(title = "Cohort Effect on TB Mortality in Sub Saharan Africa", x = "Cohort (Birth Year)", y = "Effect Size",
       caption = "TB Mortality Sub Saharan Africa(1990-2020)")
```

## Heatmap of Mortality Trends
```{r warning=FALSE, message=FALSE}
ggplot(tb_data, aes(x = Period, y = Age, fill = Mortality_Rate)) +
  geom_tile() +
  scale_fill_viridis_c() +
  theme_minimal() +
  labs(title = "Heatmap of TB Mortality Trends in SSA",
       x = "Year",
       y = "Age Group",
       fill = "Mortality Rate")
```

